Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-07-04 review_article.Rmd
CTTMB notes and added section headings

### Choropleth map (Steff)

@@

<!-- psychology -->
The familiar shapes help viewers to visually infer spatial relationships, and hence intuitive to read [@CIBMUK]. However, the different sizes of administrative areas can lead viewers to place more attention on the shades of the largest areas. The inferences derived, and the hypotheses developed can suffer from a large area "bias". <!--This technique has been seen as a 'trigger for questions' which drives research of spatial distributions of diseases.  @MACM also recognise the intuition that can be derived from choropleths with the critique that they must be 'validated by rigorous statistical analyses.' -->

<!-- Data shown -->
In epidemiology, choropleths are often used as a mapping tool in the study of the spatial distribution. 
Almost 100 years of cancer maps have explored the US and the UK with increased effectiveness in presentation of unbiased rates. 
@HEDP considered maps a research tool, dots maps use spots, choropleths and thematic maps shade areas and isopleths use isolines. These  techniques were commonly used before the advent of computer-assisted cartography.
The data collection methods of cancer mortality rates across regions, and administrative control within regions lends itself choropleth visualisation. Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the proportion of old people. Howe [-@HEDP] describes Stocks development of the standardised mortality ratios through the 1930s. The choropleth maps presented levels of cancer via hatchings on a black and white scale.

<!-- Usage and availability -->
@DMAHP cites Cruickshank's (1947) discussion of using visuals as 'formal statistical assessment of the spatial pattern' as a major advancement.
@CD considers cartographic data analysis in practice. Presenting a clear guide for preparing cartograms for use with both qualitative and quantitative information.
The authors recognise that the creation of cartograms was largely the work of professional cartographers until the innovation of geographical information systems that welcomed 
map users as map creators. These systems are utilised depending on 'the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994)'.
@SE reports the increase in disease mapping development due to the availability of geographic information system softwares.

<!--
@CPISACA
The potential audience for the results of a spatial analysis of health data is no longer limited to scientists but now also includes the public, policymakers, the media, and a host of others.

@@

<!-- Going from choropleth to cartogram -->
Cartograms are a thematic map, presenting statistical and geographical information like choropleths. @ACTUC suggests the difference is the intentional and desirable augmentation of the size, shape or distance of geographical areas. World projections reflect the frequent distortions seen from altering perspectives. Choropleth maps will always be distorted if they cover enough of the globe, as will photographs of the globe from space. Cartogram creation requires choosing a favourable distortion of the geography for presenting the set of spatial information.
This concept was originally implemented to allow a map base that represented the proportion of the population within each region. @ACTUC considers this design a 'more socially just form of mapping' by giving all members of the population 'equitable representation'. This is known as an iso-demographic map, and the design inherently lends itself to epidemiolgy. 
@HEDP argues that 'cancer occurs in people, not in geographical areas' and the map base should reflect this to avoid allocating 'undue prominence' to rural areas. @DMAHP reduce the visual impact of large areas with small populations, the distortion of spatial areas is proportional to the denominator variable.

@@

>>>>

### Current best practice map displays for cancer data
<!--Choropleth map (Steff)-->

@@

<!-- psychology -->
The familiar state shapes help viewers to visually infer spatial relationships, and hence intuitive to read [@CIBMUK]. However, the different sizes of administrative areas can lead viewers to place more attention on the shades of the largest areas. The inferences derived, and the hypotheses developed can suffer from a large area "bias". <!--This technique has been seen as a 'trigger for questions' which drives research of spatial distributions of diseases.  @MACM also recognise the intuition that can be derived from choropleths with the critique that they must be 'validated by rigorous statistical analyses.' -->

<!-- Displaying cancer data on choropleths -->
In epidemiology, choropleths are often used as a tool to study the spatial distribution of cancer incidence and mortality. 
Almost 100 years of cancer maps have explored the United States and the United Kingdom with increased effectiveness in presentation of unbiased rates. The increasing development and use of disease maps can be attributed to the availability of geographic information system softwares [@SE]. The data collection methods of cancer mortality rates across regions, and administrative control within regions lends itself choropleth visualisation. Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the proportion of old people. Howe [-@HEDP] describes Stocks development of the standardised mortality ratios through the 1930s. The choropleth maps used presented levels of cancer via hatchings on a black and white scale.

@CPISACA
The potential audience for the results of a spatial analysis of health data is no longer limited to scientists but now also includes the public, policymakers, the media, and a host of others.

<!-- Usage and availability -->
@DMAHP cites Cruickshank's (1947) discussion of using visuals as 'formal statistical assessment of the spatial pattern' as a major advancement.
@CD considers cartographic data analysis in practice. Presenting a clear guide for preparing cartograms for use with both qualitative and quantitative information.
 The authors recognise that the creation of cartograms was largely the work of professional cartographers until the innovation of geographical information systems that welcomed map users as map creators. These systems are utilised depending on ‘the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994)’.

@@

<!-- Going from choropleth to cartogram -->
<!-- Why transform?-->
Cartograms aand choropleths both present statistical and geographical information. @ACTUC suggests the difference is the intentional and desirable augmentation of the size, shape or distance of geographical areas.
Waldo Tobler [-@TVSSS] explores many graphical techniques, and suggests there are particular methods for particular purposes.

<!-- There are many alternative views of the world-->
<!-- Tobler map projections-->
Some consider choropleths true topological maps, however, if the land mass presented covers enough of the globe, there must be a transformation to visualise it in 2D.
The world projections reflect the frequent distortions seen from altering perspectives. Choropleth maps will always be distorted if they cover enough of the globe, as will photographs of the globe from space. Choropleth creation requires choosing a favourable distortion of the geography for presenting the set of spatial information.

!["Choropleth maps of the United States of America using four coordinate reference systems."](figures/ggchoroCRS.png)

<!-- Cartograms extend transformations -->
Cartograms may be seen as an extension of this concept, implementing a distorion of the geographical shape to allow a map base that represents the population within each region. 
@ACTUC considers this design a 'more socially just form of mapping' by giving all members of the population 'equitable representation'. 
@DMAHP reduce the visual impact of large areas with small populations, the distortion of spatial areas is proportional to the denominator variable. This is known as an iso-demographic map, and the design inherently lends itself to epidemiolgy.

@@

@HEDP argues that 'cancer occurs in people, not in geographical areas' and the map base should reflect this to avoid allocating 'undue prominence' to rural areas.

<!-- 
@CTTMB
choropleth maps of population characteristics exhibit serious graphic distortions when the population concerned varies greatly in density. The topological cartogram is proposed as a viable alternative in such situations

there is likely 'to manifest a growing demand for urban thematic maps'

paper presents 'the topological cartogram as a base map for the communication of the spatial variation of socio-economic data within urban areas. '

'it is possible to adjust their boundaries so that the areas which they enclose become proportional to these quantities rather than to their areal extent.'

'Firstly, that no such transformation has a unique solution, leaving the subjective choice in the hands of the individual map designer. '
Design goals:
- to maintain as much as possible of the intricate detail
- to maximize the simplicity of the transformation
- maintenance of directional accuracy
- attempt to influence and direct the map user's attention,

1) that the size of the electoral subdivisions should be made proportional to their respective electoral populations, 
2) that the contiguity relationships between the DCU's should be maintained, 
3) that true directions should be retained from the centre of the transformation, 
4) that the main shape characteristics of the DCU's should be preserved, and 
5) that the map should possess the overall shape of the built-up area. 
-->
